Mining the human microbiome for clinical insight

### 2018-12-18 abstract.tex
Eric's comments

The human microbiome is essential for health and has been implicated in many diseases.
DNA sequencing has enabled the detailed characterization of these human-associated microbial communities, leading to a rapid expansion in studies investigating the human microbiome.
In this thesis, I describe multiple projects which overcome various data analysis challenges to extract useful clinical insights from microbiome data.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes.
I leverage data collected from multiple sites per patient to identify aspiration-associated changes in the relationships between these communities, discovering new properties of the aerodigestive microbiome and suggesting new approaches for treatment.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets with standard data processing and analysis methods.
I find consistent patterns characterizing disease-associated microbiome changes and a set of shared associations which could inform clinical treatment and therapeutic development approaches for different microbiome-mediated diseases.
Enabled by this work, I also contribute to the development of a method to correct for batch effects in case-control microbiome studies.
In the fourth project, I describe a framework for rational donor selection in fecal microbiota transplant clinical trials in which knowledge derived from clinical and basic science research is used to inform which donor is selected for fecal transplants, increasing the likelihood of successful trials.
Finally, I present preliminary results analyzing the microbiome and metabolome of residential sewage as a novel platform for community-level public health surveillance.
Together, these projects demonstrate a variety of approaches to mine the human microbiome for clinically-relevant insights and suggests multiple avenues forward for translating findings from microbiome data analyses into clinical and public health impact.

>>>>

The human microbiome is essential for health and has been implicated in many diseases.
DNA sequencing has enabled the detailed characterization of these human-associated microbial communities, leading to a rapid expansion in studies investigating the human microbiome.
In this thesis, I describe multiple projects which overcome various data analysis challenges to extract useful clinical insights from microbiome data.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes.
I leverage data collected from multiple sites per patient to identify aspiration-associated changes in the relationships between these communities, discovering new properties of the aerodigestive microbiome and suggesting new approaches for treatment.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets with standard data processing and analysis methods.
I find consistent patterns characterizing disease-associated microbiome changes and a set of shared associations which could inform clinical treatment and therapeutic development approaches for different microbiome-mediated diseases.
Enabled by this work, I also contribute to the development of a method to correct for batch effects in case-control microbiome studies presented in the third project.
In the fourth project, I describe a framework for rational donor selection in fecal microbiota transplant clinical trials in which knowledge derived from clinical and basic science research is used to inform which donor is selected for fecal transplants, increasing the likelihood of successful trials.
Finally, I present preliminary results analyzing the microbiome and metabolome of residential sewage as a novel platform for community-level public health surveillance.
Together, these projects demonstrate a variety of approaches to mine the human microbiome for clinically-relevant insights and suggests multiple avenues forward for translating findings from microbiome data analyses into clinical and public health impact.

### 2018-12-18 chap1.tex
Eric's comments

The microbes that live in and on our bodies make up the human microbiome, are essential for health, and have been implicated in many diseases.
Almost all human body sites are colonized by microbes, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile \cite{sender-2016-bhratio,beck-2012-lungmicrobiome}.
These microbial communities perform essential functions for health, including fighting off and preventing infections, regulating host metabolism and interacting with the immune system, and metabolizing xenobiotics or other compounds which are indigestible by the host.
Additionally, perturbations in human microbiomes have been implicated in many diseases, including inflammatory, metabolic, neurological, and respiratory conditions \cite{beck-2012-lungmicrobiome,ibd-papa,ridaura-2013-obfmt,hsiao-2013-autism}.

@@

In this chapter, we present an approach for selecting donors in FMT trials to increase the likelihood that these trials will succeed.
In our proposed framework, a clinician applies her existing knowledge and hypotheses for how her indication of interest is being mediated by the microbiome to drive the donor selection process.
We present four types of disease models to describe underlying processes mediating microbiome-related diseases: acute dysbiosis, mediation by individual taxa, missing or overactive community function, and complex host-microbiome interactions.
We suggest associated donor selection strategies for each of these disease models, and provide case studies illustrating the process of selecting donors rationally.
Finally, I perform a power simulation which finds that most FMT trials are unlikely to be powered enough to identify the key donor bacteria mediating patient responses to FMT.
Thus, to successfully make discoveries from completed trials, clinicians should perform hypothesis-driven retrospective analyses and plan for these during the clinical trial design process, for example by collecting paired donor and patient and/or longitudinal samples.
The framework that we present encourages clinicians to leverage their clinical experience, existing microbiome research and published datasets, and the increasing availability of screened donor stools to more efficiently translate microbiome research into clinical impact.
It also suggests approaches for improving retrospective analyses of the microbiome data generated during these clinical trials.

@@

In Chapter \ref{chap:24hr}, I present preliminary analyses of untargeted metabolomics and microbiome data from residential sewage, identifying human health and activity markers with potential public health relevance.
Wastewater epidemiology has long been proposed for many applications in population health, but its implementation has so far been limited to polio surveillance \cite{polio} and illicit drug consumption monitoring \cite{Subedi2014,Ort2014}.
Standard wastewater epidemiology methods sample sewage at the wastewater treatment plant and perform targeted analyses to measure individual metabolites or viruses of interest.
Sampling at the treatment plant can capture information about large urban populations, but does not provide spatially resolved information about individual communities within the city.
Targeted analyses can confidently identify and quantify specific molecules, but cannot discover new biomarkers.
Because a major goal of public health is to reduce disparities between populations by identifying and tailoring interventions to high-risk groups \cite{Ramaswami2018,Weeramanthri2018,Khoury2016}, limited geographical resolution and measurable biomarkers makes standard wastewater epidemiology methods inappropriate for many public health applications.
Despite these limitations, adapting wastewater epidemiology and implementing it within communities could be useful for evaluating the impact of public health policies or interventions.
For example, biomarkers of human sugar consumption could be measured to directly and quickly evaluate the neighborhood-level public health impact of a citywide sugar tax, rather than requiring years to see its impact reflected in changing obesity rates.

>>>>

The microbes that live in and on our bodies make up the human microbiome and are essential for health.
Almost all human body sites are colonized by microbes, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile \cite{sender-2016-bhratio,beck-2012-lungmicrobiome}.
These microbial communities perform essential functions for health, including fighting off and preventing infections, regulating host metabolism and interacting with the immune system, and metabolizing xenobiotics or other compounds which are indigestible by the host.
Additionally, perturbations in human microbiomes have been implicated in many diseases, including inflammatory, metabolic, neurological, and respiratory conditions \cite{beck-2012-lungmicrobiome,ibd-papa,ridaura-2013-obfmt,hsiao-2013-autism}.

@@

In this chapter, I present an approach for selecting donors in FMT trials to increase the likelihood that these trials will succeed.
In this proposed framework, a clinician applies her existing knowledge and hypotheses for how her indication of interest is being mediated by the microbiome to drive the donor selection process.
I present four types of disease models to describe underlying processes mediating microbiome-related diseases: acute dysbiosis, mediation by individual taxa, missing or overactive community function, and complex host-microbiome interactions.
I suggest associated donor selection strategies for each of these disease models, and provide case studies illustrating the process of selecting donors rationally.
Finally, I perform a power simulation which finds that most FMT trials are unlikely to be powered enough to identify the key donor bacteria mediating patient responses to FMT.
Thus, to successfully make discoveries from completed trials, clinicians should perform hypothesis-driven retrospective analyses and plan for these during the clinical trial design process, for example by collecting paired donor and patient and/or longitudinal samples.
This framework encourages clinicians to leverage their clinical experience, existing microbiome research and published datasets, and the increasing availability of screened donor stools to more efficiently translate microbiome research into clinical impact.
It also suggests approaches for improving retrospective analyses of the microbiome data generated during these clinical trials.

@@

In Chapter \ref{chap:24hr}, I present preliminary analyses of untargeted metabolomics and microbiome data from residential sewage, identifying human health and activity markers with potential public health relevance.
Wastewater epidemiology has long been proposed for many applications in population health, but its implementation has so far been limited to polio surveillance \cite{polio} and illicit drug consumption monitoring \cite{Subedi2014,Ort2014}.
Standard wastewater epidemiology methods sample sewage at the wastewater treatment plant and perform targeted analyses to measure individual metabolites or viruses of interest.
Sampling at the treatment plant can capture information about large urban populations, but does not provide spatially resolved information about individual communities within the city.
Targeted analyses can confidently identify and quantify specific molecules, but cannot discover new biomarkers.
Because a major goal of public health is to reduce disparities between populations by identifying and tailoring interventions to high-risk groups \cite{Ramaswami2018,Weeramanthri2018,Khoury2016}, limited geographical resolution and measurable biomarkers makes standard wastewater epidemiology methods inappropriate for many public health applications.
Despite these limitations, adapting wastewater epidemiology and implementing it within communities could be useful for evaluating the impact of public health policies or interventions.
For example, rather than waiting years to see how a citywide sugar tax affects obesity rates, biomarkers of human sugar consumption could be measured before and after the policy roll-out to directly and quickly evaluate its impact on sugar consumption.
